---
title: "roostR workflow: detecting roosting behavior from Motus data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{roostR workflow: detecting roosting behavior from Motus data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

**roostR** converts raw Motus radio-telemetry detections into behavioral
metrics for nocturnally roosting birds. The core idea is that an animal moving
near a receiver tower produces larger changes in signal strength than a
stationary roosting bird. By tracking these signal differences over time,
roostR can detect:

- **Roost onset** — when the bird settles for the night (near sunset)
- **Roost departure** — when the bird leaves in the morning (near sunrise)
- **Restlessness** — activity spikes during the roost interval

This workflow is designed for the **per-bird-file** pattern: one dataframe
per individual, with `recvDeployName` identifying the receiver and `port`
identifying the antenna.

This vignette walks through the complete analysis pipeline using the built-in
`sparrow52550` dataset (a dark-eyed junco, *Junco hyemalis*, tag deploy
ID 52550).

### Required input

Your detection dataframe must contain at minimum:

| Column | Type | Description |
|--------|------|-------------|
| `recvDeployName` | character | Receiver deployment name (individual key within a per-bird file) |
| `port` | integer | Antenna port number |
| `time` | POSIXct UTC | Detection timestamp |
| `doy` | integer | Day of year |
| `sig` | numeric | Signal strength (dBm) |
| `Sunrise` | POSIXct UTC | Sunrise for that date and location |
| `Sunset` | POSIXct UTC | Sunset for that date and location |

Astronomical twilight columns (`AT.Start`, `AT.End`, `NT.Start`, `NT.End`,
`CT.Start`, `CT.End`) are included in `sparrow52550` and are useful for
context plots but not required by core functions.

```{r load}
library(roostR)
library(dplyr)
library(ggplot2)

data(sparrow52550)
glimpse(sparrow52550)
```

---

## Step 1: Collapse multi-antenna detections

A Motus tag can be detected on up to four antennas at the same receiver
simultaneously. `collapse_motus_time()` reduces these to one row per
receiver × port × timestamp, keeping the maximum signal (`sig`) and
computing the mean (`sig.mean`).

```{r collapse}
sp <- collapse_motus_time(sparrow52550)

cat("Rows before:", nrow(sparrow52550), "\n")
cat("Rows after: ", nrow(sp), "\n")
```

---

## Step 2: Compute activity proxies

`add_signal_diffs()` adds three columns per receiver–port combination:

- `sig.diff` — absolute change in maximum signal from the previous detection
- `sig.diff.mean` — absolute change in mean signal
- `time.diff` — minutes elapsed since the previous detection

These are the core activity proxies. Large `sig.diff` values indicate
movement; small values indicate a stationary (roosting) bird.

```{r signal-diffs}
sp <- add_signal_diffs(sp)
summary(sp$sig.diff)
```

`add_continuity_flags()` marks gaps in the detection series. With a
2-minute threshold (tuned to the ~20 second tag ping interval), detections
within 2 minutes of each other are "continuous":

```{r continuity}
sp <- add_continuity_flags(sp, threshold = 2)

# Proportion of detections preceded by a gap
mean(sp$gap, na.rm = TRUE)
```

---

## Step 3: Smooth signal differences

`add_roll_median()` computes a time-indexed centered rolling median over a
15-minute window (±7.5 min each side). This smooths micro-spikes from brief
movements while preserving genuine behavioral transitions, and correctly
handles irregular detection intervals.

```{r smooth}
sp <- add_roll_median(sp, window_min = 15)
```

---

## Step 4: Classify day and night

`add_day_night()` labels each detection as `"day"` or `"night"` based on
the per-row `Sunrise` and `Sunset` times:

```{r diel}
sp <- add_day_night(sp)
table(sp$diel)
```

Convert times to decimal hour-of-day for plotting:

```{r hours}
sp <- sp %>%
  mutate(
    hour         = lubridate::hour(time) + lubridate::minute(time) / 60,
    sunset_hour  = lubridate::hour(Sunset)  + lubridate::minute(Sunset)  / 60,
    sunrise_hour = lubridate::hour(Sunrise) + lubridate::minute(Sunrise) / 60
  )
```

A quick sanity check — rolling median should be higher during the day:

```{r plot-diel, fig.cap="Rolling median activity by diel period and day-of-year"}
ggplot(sp, aes(hour, roll_vol, color = diel)) +
  geom_point(size = 0.3, alpha = 0.4) +
  facet_wrap(~doy) +
  scale_color_manual(values = c(day = "#FDE725FF", night = "#440154FF")) +
  labs(x = "Hour (UTC)", y = "Rolling median signal difference",
       color = "Period") +
  theme_bw()
```

---

## Step 5: Detect roost timing

### Roost onset (sunset window)

`detect_roost_onset()` looks within ±90 minutes of sunset for the latest
sustained quiet period confirmed by a follow-up window. Two parameters
control confirmation: `confirm_min` (how long the quiet must persist,
60 min) and `confirm_frac` (fraction of detections that must be
low-volatility, default 0.8):

```{r roost-onset}
roost_times <- detect_roost_onset(sp,
                                  window_minutes = 90,
                                  vol_threshold  = 3,
                                  min_duration   = 10,
                                  confirm_min    = 60,
                                  confirm_frac   = 0.8)
head(roost_times)
```

### Roost departure (sunrise window)

`detect_roost_departure()` looks in the 90 minutes before through 90
minutes after sunrise for the first spike in activity:

```{r roost-departure}
leave_times <- detect_roost_departure(sp,
                                      window_before    = 90,
                                      window_after     = 90,
                                      spike_threshold  = 4,
                                      median_threshold = 3)
head(leave_times)
```

Join roost times back to the detection dataframe by night (`doy` only —
roost and departure times are per-night values shared across all receivers):

```{r add-roost-times}
sp <- add_roost_times(sp, roost_times, leave_times)
sp <- add_roost_hours(sp)
```

---

## Step 6: Overnight plotting

`wrap_hours_overnight()` shifts UTC hours so that 17:00 UTC (≈ noon EST)
becomes the left edge of the plot, keeping the full roost period visible in
one panel per night:

```{r wrap}
overnight <- wrap_hours_overnight(sp, pivot_hour = 17) |>
  mutate(night_doy = if_else(hour < 17, doy - 1L, doy))

# Split on the pivot hour so reference lines come from the correct calendar day.
# evening = doy N (sunset, roost onset); morning = doy N+1 relabelled as night N.
evening <- overnight |> filter(hour >= 17)
morning <- overnight |> filter(hour <  17)

sunset_lines  <- distinct(evening, night_doy, sunset_hour_wrap)
roost_lines   <- distinct(evening, night_doy, roost_hour_wrap)
sunrise_lines <- distinct(morning, night_doy, sunrise_hour_wrap)
leave_lines   <- distinct(morning, night_doy, leave_roost_hour_wrap)
```

```{r plot-overnight, fig.cap="Signal differences with roost onset, departure, sunrise and sunset lines"}
ggplot(overnight, aes(hour_wrap, sig.diff)) +
  geom_point(alpha = 0.4, size = 0.5) +
  geom_vline(data = sunset_lines,
             aes(xintercept = sunset_hour_wrap),
             linetype = 2, color = "#440154FF") +
  geom_vline(data = sunrise_lines,
             aes(xintercept = sunrise_hour_wrap),
             linetype = 2, color = "#95D840FF") +
  geom_vline(data = roost_lines,
             aes(xintercept = roost_hour_wrap),
             color = "#39568CFF") +
  geom_vline(data = leave_lines,
             aes(xintercept = leave_roost_hour_wrap),
             color = "#55C667FF") +
  facet_wrap(~night_doy) +
  scale_x_continuous(
    breaks = seq(15, 39, by = 3),
    labels = function(x) x %% 24
  ) +
  coord_cartesian(ylim = c(0, 50)) +
  geom_hline(yintercept = 5, color = "#FDE725FF", linetype = 2) +
  labs(x = "Hour (noon–noon EST)",
       y = "Signal difference") +
  theme_bw()
```

Purple dashed = sunset, green dashed = sunrise, blue solid = roost onset,
green solid = roost departure.

---

## Step 7: Night observation metrics

`compute_night_observation()` pairs roost onset on day N with departure on
day N+1 internally via a `doy - 1L` offset; nights with no detected departure
are excluded. It calculates, for each complete night:

- `time_roosting_hr` — total roost interval duration
- `n_bins_detected` — number of 5-minute bins with at least one detection
- `observed_time_hr` — `n_bins_detected × 5 / 60`
- `prop_time_observed` — coverage fraction

```{r night-obs}
night_metrics <- compute_night_observation(sp, bin_size_min = 5)
summary(night_metrics[c("time_roosting_hr", "observed_time_hr",
                         "prop_time_observed")])

sp <- sp |>
  select(-any_of(setdiff(names(night_metrics), "doy"))) |>
  left_join(night_metrics, by = "doy")
```

---

## Step 8: Restlessness

`calc_restless_all()` counts bouts of elevated activity (`sig.diff > 4`)
within the roost interval. Single-ping spikes are given a minimum duration
of ~22 seconds (`min_bout_min = 22/60`), matching the tag pulse interval:

```{r restless}
restless_summary <- calc_restless_all(sp,
                                      spike_threshold = 4,
                                      gap_min         = 5,
                                      min_bout_min    = 22 / 60)
sp <- sp |>
  select(-any_of(setdiff(names(restless_summary), "doy"))) |>
  left_join(restless_summary, by = "doy")
```

`add_spike_bouts()` annotates individual detections for plotting:

```{r spike-bouts}
sp <- sp |>
  select(-any_of(c("interval_start", "interval_end", "in_roost_interval",
                   "spike", "dt", "new_bout", "bout_id"))) |>
  add_spike_bouts(spike_threshold = 4, gap_min = 5)
```

```{r plot-spikes, fig.cap="Restlessness spikes (red) during the roost interval"}
ggplot(sp, aes(x = time, y = sig.diff)) +
  geom_point(color = "grey70", size = 0.5) +
  geom_point(data = filter(sp, spike), color = "red", size = 1.2) +
  geom_hline(yintercept = 4, linetype = 2, color = "blue") +
  labs(x = "Time (UTC)", y = "Signal difference") +
  theme_bw()
```

`calc_restless_rates()` normalizes by observed time so nights with low
coverage are comparable:

```{r restless-rates}
restless_rates <- calc_restless_rates(sp)
sp <- sp |>
  select(-any_of(setdiff(names(restless_rates), "doy"))) |>
  left_join(restless_rates, by = "doy")
```

---

## Step 9: Antenna bearing summary

`calc_bearing_summary()` computes the circular mean and modal antenna
bearing during each roost interval. This can reveal directional roosting
patterns relative to the receiver array:

```{r bearing}
bearing_summary <- calc_bearing_summary(sp)
bearing_summary
```

---

## Step 10: Summary plots

```{r plot-summary, fig.cap="Roost and departure timing relative to sunset and sunrise"}
# Convert to local time for interpretability
sp_est <- sp %>%
  mutate(
    roost_time_est = lubridate::with_tz(roost_time, "America/New_York"),
    leave_time_est = lubridate::with_tz(leave_roost_time, "America/New_York"),
    sunrise_est    = lubridate::with_tz(Sunrise, "America/New_York"),
    sunset_est     = lubridate::with_tz(Sunset,  "America/New_York"),
    roost_hour_est = lubridate::hour(roost_time_est) +
                       lubridate::minute(roost_time_est) / 60,
    leave_hour_est = lubridate::hour(leave_time_est) +
                       lubridate::minute(leave_time_est) / 60,
    sunrise_hour_est = lubridate::hour(sunrise_est) +
                         lubridate::minute(sunrise_est) / 60,
    sunset_hour_est  = lubridate::hour(sunset_est) +
                         lubridate::minute(sunset_est)  / 60
  )

night_est <- distinct(sp_est, doy, roost_hour_est,
                      leave_hour_est, sunrise_hour_est, sunset_hour_est,
                      prop_time_restless, restless_per_obs_hr)

ggplot(night_est, aes(doy)) +
  geom_point(aes(y = roost_hour_est)) +
  geom_line(aes(y = sunset_hour_est), linetype = "dashed") +
  labs(y = "Hour (America/New_York)", title = "Roost time vs. sunset") +
  theme_bw()

ggplot(night_est, aes(doy)) +
  geom_point(aes(y = leave_hour_est)) +
  geom_line(aes(y = sunrise_hour_est), linetype = "dashed") +
  labs(y = "Hour (America/New_York)", title = "Departure time vs. sunrise") +
  theme_bw()

ggplot(night_est, aes(doy, prop_time_restless)) +
  geom_point() +
  geom_line() +
  labs(x = "Day of year",
       y = "Proportion of observed roost time restless") +
  ylim(0, NA) +
  theme_bw()
```
