---
title: "roostR workflow: detecting roosting behavior from Motus data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{roostR workflow: detecting roosting behavior from Motus data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

**roostR** converts raw Motus radio-telemetry detections into behavioral
metrics for nocturnally roosting birds. The core idea is that an animal moving
near a receiver tower produces larger changes in signal strength than a
stationary roosting bird. By tracking these signal differences over time,
roostR can detect:

- **Roost onset** — when the bird settles for the night (near sunset)
- **Roost departure** — when the bird leaves in the morning (near sunrise)
- **Restlessness** — activity spikes during the roost interval

This vignette walks through the complete analysis pipeline using the built-in
`sparrow52550` dataset (a dark-eyed junco, *Junco hyemalis*, tag deploy
ID 52550).

### Required input

Your detection dataframe must contain at minimum:

| Column | Type | Description |
|--------|------|-------------|
| `tagDeployID` | integer | Deployment-specific tag ID (primary individual key) |
| `time` | POSIXct UTC | Detection timestamp |
| `doy` | integer | Day of year |
| `sig` | numeric | Signal strength (dBm) |
| `Sunrise` | POSIXct UTC | Sunrise for that date and location |
| `Sunset` | POSIXct UTC | Sunset for that date and location |

Astronomical twilight columns (`AT.Start`, `AT.End`, `NT.Start`, `NT.End`,
`CT.Start`, `CT.End`) are included in `sparrow52550` and are useful for
context plots but not required by core functions.

```{r load}
library(roostR)
library(dplyr)
library(ggplot2)

data(sparrow52550)
glimpse(sparrow52550)
```

---

## Step 1: Collapse multi-antenna detections

A Motus tag can be detected on up to four antennas at the same receiver
simultaneously. `collapse_motus_time()` reduces these to one row per
timestamp, keeping the maximum signal (`sig`) and computing the mean
(`sig.mean`).

```{r collapse}
sp <- collapse_motus_time(sparrow52550)

cat("Rows before:", nrow(sparrow52550), "\n")
cat("Rows after: ", nrow(sp), "\n")
```

---

## Step 2: Compute activity proxies

`add_signal_diffs()` adds three columns:

- `sig.diff` — absolute change in maximum signal from the previous detection
- `sig.diff.mean` — absolute change in mean signal
- `time.diff` — minutes elapsed since the previous detection

These are the core activity proxies. Large `sig.diff` values indicate
movement; small values indicate a stationary (roosting) bird.

```{r signal-diffs}
sp <- add_signal_diffs(sp)
summary(sp$sig.diff)
```

`add_continuity_flags()` marks gaps in the detection series. With a
2-minute threshold (tuned to the ~20 second tag ping interval), detections
within 2 minutes of each other are "continuous":

```{r continuity}
sp <- add_continuity_flags(sp, threshold = 2)

# Proportion of detections preceded by a gap
mean(sp$gap, na.rm = TRUE)
```

---

## Step 3: Smooth signal differences

`add_roll_median()` computes a centered rolling median over `k = 25`
observations (≈ 8–10 minutes). This smooths micro-spikes from brief
movements while preserving genuine behavioral transitions.

```{r smooth}
sp <- add_roll_median(sp, k = 25)
```

---

## Step 4: Classify day and night

`add_day_night()` labels each detection as `"day"` or `"night"` based on
the per-row `Sunrise` and `Sunset` times:

```{r diel}
sp <- add_day_night(sp)
table(sp$diel)
```

Convert times to decimal hour-of-day for plotting:

```{r hours}
sp <- sp %>%
  mutate(
    hour         = lubridate::hour(time) + lubridate::minute(time) / 60,
    sunset_hour  = lubridate::hour(Sunset)  + lubridate::minute(Sunset)  / 60,
    sunrise_hour = lubridate::hour(Sunrise) + lubridate::minute(Sunrise) / 60
  )
```

A quick sanity check — rolling median should be higher during the day:

```{r plot-diel, fig.cap="Rolling median activity by diel period and day-of-year"}
ggplot(sp, aes(hour, roll_vol, color = diel)) +
  geom_point(size = 0.3, alpha = 0.4) +
  facet_wrap(~doy) +
  scale_color_manual(values = c(day = "#FDE725FF", night = "#440154FF")) +
  labs(x = "Hour (UTC)", y = "Rolling median signal difference",
       color = "Period") +
  theme_bw()
```

---

## Step 5: Detect roost timing

### Roost onset (sunset window)

`detect_roost_onset()` looks within ±90 minutes of sunset for the first
sustained drop in rolling median below the threshold (3):

```{r roost-onset}
roost_times <- detect_roost_onset(sp,
                                  window_minutes = 90,
                                  vol_threshold  = 3,
                                  min_duration   = 10)
head(roost_times)
```

### Roost departure (sunrise window)

`detect_roost_departure()` looks in the 90 minutes before through 120
minutes after sunrise for the first spike in activity:

```{r roost-departure}
leave_times <- detect_roost_departure(sp,
                                      window_before    = 90,
                                      window_after     = 120,
                                      spike_threshold  = 4,
                                      median_threshold = 3)
head(leave_times)
```

Join roost times back to the detection dataframe and convert to decimal hours:

```{r add-roost-times}
sp <- add_roost_times(sp, roost_times, leave_times)
sp <- add_roost_hours(sp)
```

---

## Step 6: Overnight plotting

`wrap_hours_overnight()` shifts UTC hours so that 17:00 UTC (≈ noon EST)
becomes the left edge of the plot, keeping the full roost period visible in
one panel per night:

```{r wrap}
overnight <- wrap_hours_overnight(sp, pivot_hour = 17)

# One vertical line per DOY for each event
roost_lines   <- distinct(overnight, doy, roost_hour_wrap)
leave_lines   <- distinct(overnight, doy, leave_roost_hour_wrap)
sunset_lines  <- distinct(overnight, doy, sunset_hour_wrap)
sunrise_lines <- distinct(overnight, doy, sunrise_hour_wrap)
```

```{r plot-overnight, fig.cap="Signal differences with roost onset, departure, sunrise and sunset lines"}
ggplot(overnight, aes(hour_wrap, sig.diff)) +
  geom_point(alpha = 0.4, size = 0.5) +
  geom_vline(data = sunset_lines,
             aes(xintercept = sunset_hour_wrap),
             linetype = 2, color = "#440154FF") +
  geom_vline(data = sunrise_lines,
             aes(xintercept = sunrise_hour_wrap),
             linetype = 2, color = "#95D840FF") +
  geom_vline(data = roost_lines,
             aes(xintercept = roost_hour_wrap),
             color = "#39568CFF") +
  geom_vline(data = leave_lines,
             aes(xintercept = leave_roost_hour_wrap),
             color = "#55C667FF") +
  facet_wrap(~doy) +
  scale_x_continuous(
    breaks = seq(15, 39, by = 3),
    labels = function(x) x %% 24
  ) +
  coord_cartesian(ylim = c(0, 50)) +
  geom_hline(yintercept = 5, color = "#FDE725FF", linetype = 2) +
  labs(x = "Hour (noon–noon EST)",
       y = "Signal difference") +
  theme_bw()
```

Purple dashed = sunset, green dashed = sunrise, blue solid = roost onset,
green solid = roost departure.

---

## Step 7: Night observation metrics

`compute_night_observation()` calculates, for each individual × night:

- `time_roosting_hr` — total roost interval duration
- `observed_time_hr` — time with detections (gaps > 10 min treated as out-of-range)
- `prop_time_observed` — coverage fraction

```{r night-obs}
night_metrics <- compute_night_observation(sp, gap_threshold = 10)
summary(night_metrics[c("time_roosting_hr", "observed_time_hr",
                         "prop_time_observed")])

sp <- sp |>
  select(-any_of(setdiff(names(night_metrics), c("tagDeployID", "doy")))) |>
  left_join(night_metrics, by = c("tagDeployID", "doy"))
```

---

## Step 8: Restlessness

`calc_restless_all()` counts bouts of elevated activity (`sig.diff > 4`)
within the roost interval:

```{r restless}
restless_summary <- calc_restless_all(sp,
                                      spike_threshold = 4,
                                      gap_min = 2)
sp <- sp |>
  select(-any_of(setdiff(names(restless_summary), c("tagDeployID", "doy")))) |>
  left_join(restless_summary, by = c("tagDeployID", "doy"))
```

`add_spike_bouts()` annotates individual detections for plotting:

```{r spike-bouts}
sp <- sp |>
  select(-any_of(c("interval_start", "interval_end", "in_roost_interval",
                   "spike", "dt", "new_bout", "bout_id"))) |>
  add_spike_bouts(spike_threshold = 4, gap_min = 2)
```

```{r plot-spikes, fig.cap="Restlessness spikes (red) during the roost interval"}
ggplot(sp, aes(x = time, y = sig.diff)) +
  geom_point(color = "grey70", size = 0.5) +
  geom_point(data = filter(sp, spike), color = "red", size = 1.2) +
  geom_hline(yintercept = 4, linetype = 2, color = "blue") +
  labs(x = "Time (UTC)", y = "Signal difference") +
  theme_bw()
```

`calc_restless_rates()` normalizes by observed time so nights with low
coverage are comparable:

```{r restless-rates}
restless_rates <- calc_restless_rates(sp)
sp <- sp |>
  select(-any_of(setdiff(names(restless_rates), c("tagDeployID", "doy")))) |>
  left_join(restless_rates, by = c("tagDeployID", "doy"))
```

---

## Step 9: Summary plots

```{r plot-summary, fig.cap="Roost and departure timing relative to sunset and sunrise"}
night_plot <- sp %>%
  distinct(tagDeployID, doy, roost_hour, leave_roost_hour,
           sunset_hour, sunrise_hour, prop_time_restless,
           restless_per_obs_hr)

# Convert to EST for interpretability
sp_est <- sp %>%
  mutate(
    roost_time_est = lubridate::with_tz(roost_time, "America/New_York"),
    leave_time_est = lubridate::with_tz(leave_roost_time, "America/New_York"),
    sunrise_est    = lubridate::with_tz(Sunrise, "America/New_York"),
    sunset_est     = lubridate::with_tz(Sunset,  "America/New_York"),
    roost_hour_est = lubridate::hour(roost_time_est) +
                       lubridate::minute(roost_time_est) / 60,
    leave_hour_est = lubridate::hour(leave_time_est) +
                       lubridate::minute(leave_time_est) / 60,
    sunrise_hour_est = lubridate::hour(sunrise_est) +
                         lubridate::minute(sunrise_est) / 60,
    sunset_hour_est  = lubridate::hour(sunset_est) +
                         lubridate::minute(sunset_est)  / 60
  )

night_est <- distinct(sp_est, tagDeployID, doy, roost_hour_est,
                      leave_hour_est, sunrise_hour_est, sunset_hour_est,
                      prop_time_restless, restless_per_obs_hr)

ggplot(night_est, aes(doy)) +
  geom_point(aes(y = roost_hour_est)) +
  geom_line(aes(y = sunset_hour_est), linetype = "dashed") +
  labs(y = "Hour (America/New_York)", title = "Roost time vs. sunset") +
  theme_bw()

ggplot(night_est, aes(doy)) +
  geom_point(aes(y = leave_hour_est)) +
  geom_line(aes(y = sunrise_hour_est), linetype = "dashed") +
  labs(y = "Hour (America/New_York)", title = "Departure time vs. sunrise") +
  theme_bw()

ggplot(night_est, aes(doy, prop_time_restless)) +
  geom_point() +
  geom_line() +
  labs(x = "Day of year",
       y = "Proportion of observed roost time restless") +
  ylim(0, NA) +
  theme_bw()
```
